---
title: "Lab 06 - Ugly charts and Simpson's paradox"
author: "Lindsey Wilson"
date: "2/16/23"
output: github_document
---

### Load packages and data

```{r load-packages, message = FALSE}
library(tidyverse) 
library(dsbox)
library(mosaicData) 
```

```{r load-data}
staff <- read_csv("data/instructional-staff.csv")
```
### Instructional Staff Employment Trends

Here's our code to take the data in `staff` and make it long:

```{r staff-data-longer}
staff_long <- staff %>%
  pivot_longer(cols = -faculty_type, names_to = "year") %>%
  mutate(value = as.numeric(value))

staff_long
```

This allows us to plot our data as a line graph:

```{r staff-data-line-graph}
staff_long %>%
  ggplot(aes(x = year, y = value, color = faculty_type)) +
  geom_line()
```

However, when we run that code we get an error message. We can fix this by grouping by `faculty_type` in ggplot:

```{r staff-data-line-graph-fixed}
staff_long %>%
  ggplot(aes(x = year,
             y = value,
             group = faculty_type,
             color = faculty_type)) +
  geom_line() + 
  labs(title = "Hires by Faculty Type over Time",
       x = "Year",
       y = "Percentage")
```

If I wanted to show that the percentage of part-time faculty in particular has gone up, I would probably filter the data just to show that one group.

### Fisheries

```{r read-fish-data}
fisheries <- read_csv("data/fisheries.csv")
```

Let's try to plot this data in a way that isn't horrible:

```{r fish-plot}
ggplot(data = fisheries,
       mapping = aes())




```


### Exercise 1

Remove this text, and add your answer for Exercise 1 here.
Add code chunks as needed.
Don't forget to label your code chunk.
Do not use spaces in code chunk labels.

### Exercise 2

Remove this text, and add your answer for Exercise 1 here.
Add code chunks as needed.
Don't forget to label your code chunk.
Do not use spaces in code chunk labels.

### Exercise 3

...

Add exercise headings as needed.
